Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2017-05-24 paper.tex
...

The rest of the paper is organized as follows:
Section~\ref{sec.codebase} gives an overview of the Pingus codebases in C++ and
\CEU and describes our approach to identify and rewrite the control flow in the
game.
Section~\ref{sec.patterns} discusses six case studies in detail which are
categorized in five control-flow patterns.
Section~\ref{sec.related} discusses related work.
Section~\ref{sec.conclusion} concludes the paper.

@@

\section{Control-Flow Patterns \& Case Studies}
\label{sec.patterns}

@@

TODO: state pattern, FSM
      strategy/visitor pattern, polymorhpism, hierarchies
      observer pattern, signaling

\subsection{Finite State Machines}

@@

\subsection{Continuation Passing}

@@

\subsection{Dispatching Hierarchies}

@@

\subsection{Lifespan Hierarchies}

@@

In C++, the class \code{PinguHolder} is a container that holds all pingus
alive.
%
The method \code{PinguHolder::create\_pingu} (ln. 1--6) is called periodically
to create a new \code{Pingu} and add it to the \code{pingus} collection
(ln. 3--4).
The method \code{PinguHolder::update} (ln. 8--18) checks the state of all
pingus on every frame, removing those with the dead status (ln. 12--14).
%
Entities with dynamic lifespan in C++ require explicit \code{add} and
\code{remove} calls associated to a container (ln. 4,13).
Without the \code{erase} call above, a dead pingu would remain in the
collection with updates on every frame (ln. 11).
Since the \code{redraw} behavior for a dead pingu is innocuous, the death could
go unnoticed but the program would keep consuming memory and CPU time.
This problem is known as the \emph{lapsed listener}~\cite{games.patters} and
also occurs in languages with garbage collection:
A container typically holds a strong reference to a child (sometimes the only 
reference to it), and the runtime cannot magically detect it as garbage.

@@

\subsection{Signaling Mechanisms}

@@

In \CEU, a \emph{Ctrl-G} key press broadcasts the internal event
\code{config\_manager.go\_grab} to the application (ln. 8).
%
The configuration manager just needs to react to \code{go\_grab}
continuously to perform the \emph{grab} effect (ln. 25--27).
%
The \code{ChkBox} exposes the event \code{go\_click} for notifications in
both directions, i.e., from the abstraction to the application and
\emph{vice versa}:
%
The abstraction reacts to external clicks continuously (ln. 41--43) to
broadcast the event \code{go\_click} (ln. 47).
It also reacts continuously to \code{go\_click} in another line of execution
(ln. 45--48), which awakes from notifications from the first line of execution
or from the application.
%
The \code{OptMenu} connects the two events as follows:
%
The two loops in parallel handle the connections in opposite directions:
from the configuration manager to the check box (ln. 60--62);
and
from the check box to the configuration manager (ln. 65--67).
%
When the \emph{Option screen} terminates, the connections break automatically
since the body with the two loops is automatically aborted.
%
Note that the implementation in \CEU does not check event emits to break the
dependency cycle and prevent infinite execution.
Due to the stack-based execution for internal events in
\CEU~\cite{ceu.sensys13)}, programs with mutually-dependent events do not
create infinite execution loops.

@@

The control-flow patterns closely relate to the \emph{GoF} behavioral
patterns~\cite{gof}, which some previous work discuss in the context of video
game development~\cite{a,b,c}.
%
different
    - generally?
        - mostram exemplos abstratos
            - imagine um jogo...
                - assim poderiamos aplicar...
            - sem codigo real
            - nao aplicam a jogos existentes
    - propose solutions
        - eliminate
for instance, the \emph{state pattern}, state machines, sequences
trivial case of the bomber animation is a series of statements separated
by await statements
visitor, adopted in the language

>>>>

The rest of the paper is organized as follows:
Section~\ref{sec.codebase} gives an overview of the Pingus codebases in C++ and
\CEU and describes our approach to identify and rewrite the control flow in the
game.
Section~\ref{sec.pats} discusses six case studies in detail which are
categorized in five control-flow patterns.
Section~\ref{sec.related} discusses related work.
Section~\ref{sec.conclusion} concludes the paper.

@@

\section{Control-Flow Patterns \& Case Studies}
\label{sec.pats}

@@

\subsection{Finite State Machines}
\label{sec.pats.fsms}

@@

\subsection{Continuation Passing}
\label{sec.pats.cps}

@@

\subsection{Dispatching Hierarchies}
\label{sec.pats.dispatching}

@@

\subsection{Lifespan Hierarchies}
\label{sec.pats.lifespan}

@@

In C++, the class \code{PinguHolder} is a container that holds all pingus
alive.
%
The method \code{PinguHolder::create\_pingu} (ln. 1--6) is called periodically
to create a new \code{Pingu} and add it to the \code{pingus} collection
(ln. 3--4).
The method \code{PinguHolder::update} (ln. 8--18) checks the state of all
pingus on every frame, removing those with the dead status (ln. 12--14).
%
Entities with dynamic lifespan in C++ require explicit \code{add} and
\code{remove} calls associated to a container (ln. 4,13).
Without the \code{erase} call above, a dead pingu would remain in the
collection with updates on every frame (ln. 11).
Since the \code{redraw} behavior for a dead pingu is innocuous, the death could
go unnoticed but the program would keep consuming memory and CPU time.
This problem is known as the \emph{lapsed listener}~\cite{games.patterns} and
also occurs in languages with garbage collection:
A container typically holds a strong reference to a child (sometimes the only 
reference to it), and the runtime cannot magically detect it as garbage.

@@

\subsection{Signaling Mechanisms}
\label{sec.pats.signaling}

@@

In \CEU, a \emph{Ctrl-G} key press broadcasts the internal event
\code{config\_manager.go\_grab} to the application (ln. 8).
%
The configuration manager just needs to react to \code{go\_grab}
continuously to perform the \emph{grab} effect (ln. 25--27).
%
The \code{ChkBox} exposes the event \code{go\_click} for notifications in
both directions, i.e., from the abstraction to the application and
\emph{vice versa}:
%
The abstraction reacts to external clicks continuously (ln. 41--43) to
broadcast the event \code{go\_click} (ln. 47).
It also reacts continuously to \code{go\_click} in another line of execution
(ln. 45--48), which awakes from notifications from the first line of execution
or from the application.
%
The \code{OptMenu} connects the two events as follows:
%
The two loops in parallel handle the connections in opposite directions:
from the configuration manager to the check box (ln. 60--62);
and
from the check box to the configuration manager (ln. 65--67).
%
When the \emph{Option screen} terminates, the connections break automatically
since the body with the two loops is automatically aborted.
%
Note that the implementation in \CEU does not check event emits to break the
dependency cycle and prevent infinite execution.
Due to the stack-based execution for internal events in
\CEU~\cite{ceu.sensys13}, programs with mutually-dependent events do not
create infinite execution loops.

@@

The control-flow patterns closely relate to the \emph{GoF} behavioral
patterns~\cite{gof}, which some previous work discuss in the context of video
game development~\cite{games.patterns,games.gof.2015,games.gof.2007}.
%
The original game in C++ uses variations of the
    \emph{state} (Sections~\ref{sec.pats.fsms}~and~\ref{sec.pats.cps}),
    \emph{visitor} (Sections~\ref{sec.pats.dispatching}~and~\ref{sec.pats.lifespan}), and
    \emph{observer} (Section~\ref{sec.pats.signaling} and to handle input in general)
patterns as implementation details to achieve the desired higher-level
control-flow patterns.
%
\CEU overcomes the need of behavioral patterns with a semantics that supports
structured control-flow mechanisms and event-based communication via broadcast.
%
As an example, the \emph{state pattern} for the bomber animation in
Section~\ref{sec.pats.fsms} becomes a series of blocks separated by
\code{await} statements.

A number of domain-specific languages, frameworks, and techniques have been
proposed for particular subsystems of the game logic, such as
animations~\cite{games.anims.2006,games.anims.2003,games.anims.1996,games.anims.1982},
game state and screen progression~\cite{games.fsms.2006.1,games.fsms.2006.2}, and
behavior and AI modeling~\cite{games.bts,games.bts.unreal}
%
In Pingus, we employed \CEU at the core of the game for event dispatching
(Section~\ref{sec.pats.dispatching}) and memory management of entities
(Section~\ref{sec.pats.lifespan}), eliminating parts of the original game
engine.
We also implemented all entity animations and behaviors
(Section~\ref{sec.pats.fsms}), screen transitions
(Section~\ref{sec.pats.cps}), and intermodule communication
(Section~\ref{sec.pats.signaling}) using the available control mechanisms.
%
\CEU is a superset of C targeting reactive systems in general, not only games,
and has also been adopted in other domains, such as
    wireless sensor networks~\cite{ceu.sensys13,ceu.terra} and
    multimedia systems~\cite{ceu.media.webmedia16}.

Functional reactive programming is an
FRP
    - interfacing with C
    - live porting

different
    - generally?
        - mostram exemplos abstratos
            - imagine um jogo...
                - assim poderiamos aplicar...
            - sem codigo real
            - nao aplicam a jogos existentes
    - propose solutions
        - eliminate
visitor, adopted in the language
